---
title: "Forecasting Competition"
subtitle: "Assignment A08"
author: "Gaby Czarniak & Weilin Wang"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# GitHub Repository: 
The following repository is private and Luana (lmmlima) has been invited as a collaborator: https://github.com/gaby-cz/CzarniakWang_ENV797_TSA_ForecastCompetition_S25.git 


```{r Loading libraries, message=FALSE, warning=FALSE, include=FALSE}

library(readxl)
library(dplyr)
library(tidyr)
library(forecast)
library(lubridate)
library(ggplot2)
library(zoo)
library(cowplot)
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
```
# Code for importing data set

```{r Importing data, echo=TRUE, message=FALSE, warning=FALSE}

# Importing time series data from Excel files
load_data <- read_excel("datasets/load.xlsx")
humidity_data <- read_excel("datasets/relative_humidity.xlsx")
temperature_data <- read_excel("datasets/temperature.xlsx")

```

# Code used to wrangle the data set

## Inspecting the data

```{r Inspect data}

# Inspecting the data
head(load_data)
head(humidity_data)
head(temperature_data)

```

## Addressing missing values 

```{r Handling missing values}

summary(load_data)
# We have 6 NAs in h2

summary(humidity_data)
# No NAs in humidity data

summary(temperature_data)
# No NAs in temperature data

# Identify NAs in load data
NAs <- load_data %>% 
  filter(is.na(load_data$h2) == TRUE)
NAs
# 5 of the NAs are between 2005-2010

# Getting number of variables and observations for load
nvar <- ncol(load_data) - 1
nobs <- nrow(load_data)

# Interpolate missing values for load
load_data <- load_data %>%
        mutate(h2 = na.approx(h2))

NAcheck <- load_data %>% 
  filter(is.na(load_data$h2) == TRUE)
NAcheck

```

## Creating individual and joint data frames

```{r Process the data}

# Take the mean load from h1 through h24 for each day
load_daily <- load_data %>%
  rowwise() %>%
  mutate(Daily_Load = mean(c_across(h1:h24))) %>%
  select(date, Daily_Load)

# Take the daily averages of humidity data across sensors
humidity_daily <- humidity_data %>%
  group_by(date) %>%
  summarize(Daily_Humidity = mean(c_across(starts_with("rh_ws")), na.rm = TRUE))

# Take the daily averages of temperature data across sensors
temperature_daily <- temperature_data %>%
  group_by(date) %>%
  summarize(Daily_Temperature = mean(c_across(starts_with("t_ws")), na.rm = TRUE))

# Join load, humidity, and temperature variables into one data frame
daily_data <- load_daily %>%
  left_join(humidity_daily, by = "date") %>%
  left_join(temperature_daily, by = "date")

```

### Summary statistics were inspected to assess need for log-transformations before calculating correlations. See .rmd for this code.


```{r Inspect summary statistics for load, eval=FALSE, include=FALSE}

head(daily_data)

summary(daily_data$Daily_Load)

summary(daily_data$Daily_Humidity)

summary(daily_data$Daily_Temperature)

# It doesn't look like we need to log-transform any variables before calculating correlation

```

## Inspecting correlation among variables

```{r Inspect correlation among variables}

# Create one ts object for all columns
all_ts <- msts(daily_data[,2:4], 
             seasonal.periods = c(7, 365), 
             start = c(2005, 1, 1))
plot(all_ts)

# Calculate correlation matrix R and Cholesky decomposition R
R = cor(all_ts)
print(R)

# Temperature and humidity are fairly correlated, at R = 0.38. Load and temperature are negatively correlated, albeit more weakly, at R = -0.26. Load and humidity are more weakly correlated at R = -0.19. 

```

# Creating a time series object using multiple seasonality

```{r Creating a load-only ts object}
# Creating our first ts object
# Using msts because our daily data has more than one seasonal component
daily_ts <- msts(daily_data$Daily_Load, 
                 seasonal.periods = c(7, 365), 
                 # start year, month, day
                 start = c(2005, 1, 1))

# Inspect the ts plot
autoplot(daily_ts) + ggtitle("Daily Load Time Series")

```

## Decomposing the load series

```{r Decompose the series}

# Since we have multiple seasonality, we need to use mstl() for decomposition
decompose_load <- mstl(daily_ts)
autoplot(decompose_load)
# Pretty cool!

```

## Creating subsets for training, testing, and forecasting purposes

```{r Create subsets for training and testing purposes}

# Subset for training
# For model training, use data from January 1, 2005, to December 31, 2009.
n_for = 365

daily_ts_train <- subset(daily_ts,
                         end = length(daily_ts) - n_for)
autoplot(daily_ts_train) + ggtitle("Training subset")

# Subset for testing -- can delete if we don't end up needing
daily_ts_test <- subset(daily_ts,
                        start = length(daily_ts) - n_for)
autoplot(daily_ts_test) + ggtitle("Testing subset")

# To evaluate model performance, use the fitted models to forecast daily demand for January 1, 2010, to February 28, 2010, as a validation set.

```


# Model 1: Our top-performing model on Kaggle

```{r SARIMA autofit for load only}

# Gaby added
# Simple SARIMA autofit on original series

SARIMA_autofit <- auto.arima(daily_ts[1:(5 * 365)])
print(SARIMA_autofit)

plot_grid(
  autoplot(SARIMA_autofit$residuals),
  autoplot(Acf(SARIMA_autofit$residuals,
                lag.max = 40,
                plot = FALSE),
           main = ""),
  autoplot(Pacf(SARIMA_autofit$residuals,
                lag.max = 40,
                plot = FALSE),
           main = ""),
  nrow = 1,
  ncol = 3
)

SARIMA_forecast <- forecast(SARIMA_autofit, h = 59)
print(SARIMA_forecast)

```

```{r Write SARIMA to csv, eval=FALSE, include=FALSE}

write.csv(SARIMA_forecast, "SARIMA_forecast_1_T1.csv") 
# This model got a public score of 22.29 -- KEEP

```

# Model 2: ARIMA with Fourier terms c(2,2)

```{r ARIMA with Fourier terms 22}

# Fourier terms as exogenous regressors
# Fourier terms are ideal for complex seasonalities, like in the load data (both daily and hourly variation)
# P, D, and Q are 0 because seasonality will be expressed in the Fourier terms

# First, store the Fourier terms K values consistently
K_values <- c(2, 2) 

# Fit the model with these K values
ARIMA_Four_fit <- auto.arima(daily_ts_train,
                           seasonal = FALSE,
                           lambda = 0,
                           xreg = fourier(daily_ts_train, K = K_values))

# Create Fourier terms for forecasting with the same K values
fourier_terms <- fourier(daily_ts_train, K = K_values, h = 59)

# Generate forecast using the model
ARIMA_Four_forecast <- forecast(ARIMA_Four_fit,
                              xreg = fourier_terms,
                              h = 59)  # 59 days
```

# Model 3: ARIMA with Fourier terms c(2,4)

```{r Fourier 24}
# First, store the Fourier terms K values consistently
K_values <- c(2, 2) 

# Fit the model with these K values
ARIMA_Four_fit <- auto.arima(daily_ts_train,
                           seasonal = FALSE,
                           lambda = 0,
                           xreg = fourier(daily_ts_train, K = K-values))

# Create Fourier terms for forecasting with the same K values
fourier_terms <- fourier(daily_ts_train, K = K_values, h = 59)

# Generate forecast using the model
ARIMA_Four_forecast <- forecast(ARIMA_Four_fit,
                              xreg = fourier_terms,
                              h = 59)  # 59 days

```

# Model 4: ETS + STL

```{r ETS + STL}

## ETS needs to be applied to the deseasoned series. Then, we add the seasonal component forecast back in to get the final forecast. 

# Fit and forecast STL + ETS model to data -- Gaby added
ETS_fit <- stlf(daily_ts_train, h = 59)

# Plot forecasting results -- Gaby added
autoplot(ETS_fit) + ylab("Load")

# Plot model + observed data -- Gaby added
autoplot(daily_ts) +
  autolayer(ETS_fit, series = "STL + ETS", PI = FALSE) +
  ylab("Load")

```

```{r Save ETS STL updated, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

write.csv(ETS_fit, "ETSSTL_forecast_1_T1.csv") 

# This model got a private score of 29.8
```

# Model 5: ARIMA with Fourier terms c(2,12)

```{r Fourier 212}
# First, store the Fourier terms K values consistently
K_values <- c(2, 12) 

# Fit the model with these K values
ARIMA_Four_fit <- auto.arima(daily_ts_train,
                           seasonal = FALSE,
                           lambda = 0,
                           xreg = fourier(daily_ts_train, K = K-values))

# Create Fourier terms for forecasting with the same K values
fourier_terms <- fourier(daily_ts_train, K = K_values, h = 59)

# Generate forecast using the model
ARIMA_Four_forecast <- forecast(ARIMA_Four_fit,
                              xreg = fourier_terms,
                              h = 59)  # 59 days

```


The five models submitted that performed best among this group were, in order from best to worst: SARIMA, ARIMA with Fourier terms c(2,2), ARIMA with Fourier terms c(2,4), ETS + STL, and ARIMA with Fourier terms c(2,12). In deciding which models to include, performance was judged based on Kaggle Private Score. Had Public Score been utilized, the seasonal naive on the original data would have bumped ETS + STL to fifth place and ARIMA with Fourier terms c(2,12) would have fallen off the list of best five performing models.


## See .rmd file for other files attempted.


```{r Arithmetic mean on original data, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Gaby added
# Train on January 1, 2005, to December 31, 2009.
MEAN_seas_load_train <- meanf(y = daily_ts_train, h = 59)
plot(MEAN_seas_load_train)

# After selecting the best-performing model(s), re-train the model using the full dataset from January 1, 2005, to December 31, 2010. Use this updated model to generate forecasts for January 1, 2011, to February 28, 2011. The final forecasts will be submitted as part of the competition.
MEAN_seas_load_forecast <- meanf(y = daily_ts, h = 59)
plot(MEAN_seas_load_forecast)

```

```{r Seasonal naive on original data, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Gaby added
# Seasonal naive on original data

SNAIVE_seas_train <- snaive(daily_ts_train, h=59)
plot(SNAIVE_seas_train)

SNAIVE_seas_forecast <- snaive(daily_ts, h=59)
plot(SNAIVE_seas_forecast)

write.csv(SNAIVE_seas_forecast, "forecast1_T1.csv")

```

```{r Simple moving average on original data, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Gaby added
# Simple moving average on original data

SMA_seas_fit_train <- sma(y = daily_ts_train, h = 59, 
                          holdout = FALSE) # silent = FALSE gives an error
summary(SMA_seas_fit_train)

SMA_seas_forecast_train <- forecast(object = SMA_seas_fit_train, h = 59)
plot(SMA_seas_forecast_train)

SMA_seas_fit <- sma(y = daily_ts, h = 59, 
                          holdout = FALSE) # silent = FALSE gives an error

SMA_seas_fit_forecast <- forecast(object = SMA_seas_fit, h = 59)
plot(SMA_seas_fit_forecast)

SMA_seas_fit_forecast_summary <- summary(SMA_seas_fit_forecast)

write.csv(SMA_seas_fit_forecast, "SMA_seas_fit_1_T1.csv")

```

```{r SES model, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
## Could add an exponential smoothing with ses(), arithmetic mean with meanf(), and could experiment with non-seasonal naive on the deseasoned series with naive()

# Running the below on the training dataset + h gave very off values
SES_fit_train <- ses(daily_ts_train, h=59)
summary(SES_fit_train)
autoplot(SES_fit_train)+
  autolayer(daily_ts_test) # not a great model!

SES_forecast <- ses(daily_ts, h=59)
plot(SES_forecast)

write.csv(SES_fit, "SES_forecast_1_T1.csv") ## DROP
```

```{r Non-seasonal naive, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#Creating non-seasonal ts 
deseasonal_load <- seasadj(decompose_load)

# Naive on deseasonal load data
NAIVE_deseas_load <- naive(deseasonal_load, h = 59)
plot(NAIVE_deseas_load)

```


```{r Performance metrics for our own trial and error, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Weilin add
# 1. Calculate MAPE for all models on the validation period
actual <- daily_ts[(5 * 365 + 1):(5 * 365 + 59)]

mape_snaive <- mean(abs((actual - SNAIVE_seas_train$mean) / actual)) * 100
mape_sma <- mean(abs((actual - SMA_seas_fit_train$mean) / actual)) * 100
mape_arima <- mean(abs((actual - arima_forecast$mean) / actual)) * 100
mape_ets <- mean(abs((actual - ets_forecast$mean) / actual)) * 100
mape_tbats <- mean(abs((actual - tbats_forecast$mean) / actual)) * 100

mape_results <- data.frame(
  Model = c("Seasonal Naive", "Simple Moving Average", "ARIMA", "ETS", "TBATS"),
  MAPE = c(mape_snaive, mape_sma, mape_arima, mape_ets, mape_tbats)
)

mape_results <- mape_results[order(mape_results$MAPE), ]
print(mape_results)

# 2. Train the best model on the full dataset and generate forecasts
best_model <- tbats(daily_ts)
best_forecast <- forecast(best_model, h = 59)

# 3. Format for submission
submission_df <- data.frame(
  date = format(seq.Date(from = as.Date("2011-01-01"), 
                   by = "day", 
                   length.out = 59), "%Y-%m-%d"),
  load = round(best_forecast$mean)
)

write.csv(submission_df, "final_submission.csv", row.names = FALSE)

```

```{r Fitting SARIMA to temp and humidity, eval=FALSE, include=FALSE}

plot(all_ts[,2], main = "Daily Humidity")
humidity_SARIMA <- auto.arima(all_ts[(1:(5 * 365)),2], seasonal=TRUE)
summary(humidity_SARIMA)

plot(all_ts[,3], main = "Daily Temperature")
temp_SARIMA <- auto.arima(all_ts[(1:(5 * 365)),3], seasonal = TRUE)
summary(temp_SARIMA)

```

```{r ARIMA on deseasoned, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Hi Weilin! Since we have a SARIMA, I'm going to run this ARIMA with seasonal = FALSE because otherwise we try the same thing twice. We can use the nonseasonal ARIMA on deseasoned and then add in a model of just the seasonal component and see how it works.

arima_deseas_load <- auto.arima(deseasonal_load, seasonal = FALSE)
arima_model_deseas_load

arima_deaseas_load_forecast <- forecast(arima_deseas_load, seasonal = FALSE)
plot(arima_deaseas_load_forecast)

```

```{r ETS STL attempt dropped, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# ------------------------------

ets_model <- ets(daily_ts[1:(5 * 365)])  # Training on first 5 years
ets_forecast <- forecast(ets_model, h = 59)

write.csv(ets_forecast, "ETS_forecast_1_T1.csv") ## This model got a public score of 83.96 -- DROP

```

```{r TBATS, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

tbats_model <- tbats(daily_ts[1:(5 * 365)])
tbats_forecast <- forecast(tbats_model, h = 59)

```
